### 2023-02-17 chapters/introduction.tex
connecting chapters

\chapter{Introduction}
The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, education~\cite{luckin2018machine}, and transportation~\cite{peled2019model,zantalis2019review}. In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users~\cite{eshete2021making}. With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox}, and privacy~\cite{papernot2016towards} in predictions. In this thesis, we focus on the interpretability and fairness aspects of machine learning. The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.

Towards responsible and trustworthy machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers. \emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features. Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI). In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers. In another research theme of \emph{fairness in machine learning}, unregulated classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless classifiers are trained with a fairness objective. Consequently, research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms. In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features. Finally, we explain the sources of bias based on fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier. To summarize our thesis on the interpretability and fairness in machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.

@@

The problem of learning rule-based classifiers is known to be computationally intractable. The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}. In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc. In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}. Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size. The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets. Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP (mixed integer linear programming) to efficiently learn rule-based classifiers in a mini-batch learning setting.  Here we summarize the contributions of the thesis on interpretable rule-based machine learning.

\subsection*{Scalability via Incremental Learning}
We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for learning interpretable classification rules expressed in CNF. $ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query. Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to real-world classification datasets of thousands to millions of samples. Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning. The resulting framework learns a classifier by iteratively covering the training data, wherein in each iteration, it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch. In our experiments, $ \mathsf{IMLI} $ achieves the best balance among prediction accuracy, interpretability, and scalability. For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability with respect to existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail. As an application, we deploy $ \mathsf{IMLI} $ in learning other interpretable representations: DNF classifiers, decision lists, and decision sets.

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}. Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule. Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and the higher expressiveness of Boolean cardinality constraints. Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets. While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive. To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules. Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.

@@

In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers. Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas. {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}. Because SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby extending the verification problem to multiple sensitive features unlike earlier methods. In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic} because of the efficient SSAT encoding.
%, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}. 
%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.

@@

\subsection*{Explaining Fairness metrics: Identifying the Sources of Bias}
Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}. In this thesis, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.} Building on global variance decomposition, we propose an algorithm {\fairXplainer} to estimate the fairness influence function (FIF) of features. In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods. We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.

\section{Tools}
	We develop following open-source tools for interpretable and fair machine learning.
	\begin{itemize}
		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
		\item \red{FairXplainer for explaining group-based fairness metrics in machine learning.}
	\end{itemize}
\section{Outline}
We organize the thesis in the following way. We discuss preliminaries in Chapter~\ref{chapter:preliminaries}. In the first part of the thesis on interpretable machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers. We conclude this part by applying incremental learning on a more expressible yet interpretable rule-based classifier in Chapter~\ref{chapter:crr}. In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. We conclude this part by explaining group fairness metrics in Chapter~\ref{chapter:fif}. Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.

>>>>

\chapter{Introduction}
The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, education~\cite{luckin2018machine}, and transportation~\cite{peled2019model,zantalis2019review}. In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users~\cite{eshete2021making}. With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox}, and privacy~\cite{papernot2016towards} in predictions. In this thesis, we focus on the interpretability and fairness aspects of machine learning. The thesis establishes a close integration of formal methods and automated reasoning with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.

Towards responsible and trustworthy machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers. \emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features. Interpretable rule-based classifiers not only interpret the decision function but also are applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI). In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers. In another research theme of \emph{fairness in machine learning}, unregulated classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless classifiers are trained with a fairness objective. Consequently, research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms. In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features. Finally, we explain the sources of bias by formalizing and computing fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier. To summarize our thesis on interpretability and fairness in machine learning, we prioritize on improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.

@@

The problem of learning rule-based classifiers is known to be computationally intractable. The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}. In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc. In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}. Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size. The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets. Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP to efficiently learn rule-based classifiers in a mini-batch learning setting.  Here we summarize the contributions of the thesis on interpretable rule-based machine learning.

\subsection*{Scalability via Incremental Learning}
We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for learning interpretable classification rules expressed in CNF. $ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and sparsity of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query. Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to real-world classification datasets of thousands to millions of samples. Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning. The resulting framework learns a classifier by iteratively covering the training data, wherein in each iteration, it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch. In our experiments, $ \mathsf{IMLI} $ achieves the best balance among prediction accuracy, interpretability, and scalability. For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability with respect to existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail. As an application, we deploy $ \mathsf{IMLI} $ in learning other interpretable representations: DNF classifiers, decision lists, and decision sets.

\subsection*{Expressiveness via Logical Relaxation}
We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness. Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule. Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and the higher expressiveness of Boolean cardinality constraints. Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets. While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive. To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules. Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.

@@

In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers. Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas. {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}. SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby it is natural in SSAT-based formulation to extend the verification problem to multiple sensitive features unlike earlier methods. In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic} because of the efficient SSAT encoding.
%, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}. 
%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.

@@

\subsection*{Explaining Fairness Metrics: Identifying Sources of Bias}
Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias. Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}. In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset. To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier. Thus, \textit{we aim to design a bias attribution method that is global by construction}. In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features. Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.  In our contribution, we propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features. In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods. We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.

\section{Tools}
	We develop following open-source tools for interpretable and fair machine learning.
	\begin{itemize}
		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
		\item \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for explaining group-based fairness metrics in machine learning.
	\end{itemize}
\section{Outline}
We organize the thesis in the following way. We discuss preliminaries in Chapter~\ref{chapter:preliminaries}. In the first part of the thesis on interpretable machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers. We conclude this part by applying incremental learning on a more expressible yet interpretable rule-based classifier in Chapter~\ref{chapter:crr}. In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}. We conclude this part by explaining group fairness metrics in Chapter~\ref{chapter:fif}. Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.
